---
title: 'Chapter 03: Individual geoms'
author: "Stan Piotrowski"
date: "`r format(Sys.Date(), '%B %d %Y')`"
output:
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r warning = FALSE, message = FALSE}
# Libraries
library(tidyverse)
library(patchwork)
library(kableExtra)
library(ggtext)
library(ggrepel)

# ggplot2 theme
my_theme <- theme(
  panel.grid = element_blank(),
  panel.background = element_rect(fill = "white", color = "black"),
  axis.title = element_markdown(),
  plot.title = element_markdown(),
  plot.subtitle = element_markdown()
)
```


## Introduction

This chapter focuses on exploring individual geoms: geoms which take `x` and `y` aesthetics, as well as color, size, and fill.  

## Exercises

1) _What geoms would you use to draw each of the following named plots?_

For each of the following plots, we'll use the `starwars` dataset from dplyr, which contains data on different characters in the saga including name, height, mass, home planet, the films they appear in, species, and other information.

```{r}
glimpse(starwars)
```

    1.1) _Scatterplot_
    
  For a scatterplot, we would use the `goem_point()` geom.  Let's use a scatterplot to explore the relationship between the height (cm) and mass (kg) of different Star Wars characters.  Note, we'll remove any entries with missing data in either of these fields before plotting.
    
```{r}
starwars %>% 
  filter(!is.na(height) & !is.na(mass)) %>% 
  ggplot(aes(height, mass)) +
  geom_point() +
  geom_label_repel(aes(label = name), 
                   data = starwars %>% filter(grepl("Jabba", name, ignore.case = TRUE))) +
  geom_label_repel(aes(label = name), 
                   data = starwars %>% arrange(mass) %>% head(n = 1)) +
  my_theme +
  labs(x = "Height (cm)", 
       y = "Mass (kg)", 
       title = "Relationship between height and mass of Star Wars characters", 
       subtitle = "Heaviest and lightest characters by mass are labeled")
```

    1.2) _Line chart_
    
  To create a line chart to connect observations in the order of the variable on the x-axis, we would use the `geom_line()` geom. 

    1.3) _Histogram_
    
  To create a histogram, we would use the `geom_histogram()` geom.  Alternatively, we could use the `geom_bar()` geom and use a statistical transformation to bin and count observations.  For example, we can visualize the distribution of mass across sexes.
    
```{r}
starwars %>% 
  filter(!is.na(sex) & !is.na(mass)) %>% 
  ggplot(aes(mass)) +
  geom_histogram(aes(fill = sex), bins = 50) +
  facet_wrap(~sex) +
  labs(x = "Mass (kg)", 
       y = "Count", 
       title = "Distribution of mass (kg) across sexes in Star Wars characters") +
  my_theme +
  theme(strip.background = element_rect(fill = "lightgrey", color = "black"), 
        legend.position = "none")
```
    

    1.4) _Bar chart_

  To create a bar chart, we would use the `geom_bar()` geom with the `stat = "identity"` argument to ensure that the data are plotted "as is" without statistical transformation.  We can create a bar chart of the number of different characters by species overall.  
  
```{r}
starwars %>% 
  group_by(species) %>% 
  summarise(n = n()) %>% 
  filter(!is.na(species) & n > 1) %>% 
  ggplot(aes(fct_reorder(species, n, .desc = TRUE), n)) +
  geom_bar(stat = "identity") +
  labs(x = "Species", 
       y = "n", 
       title = "Number of occurrences of each species across Star Wars films") +
  my_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
  

    1.5) _Pie chart_
    
  To draw a pie chart, we'll use the `geom_bar()` geom with `stat = "identity"` argument and use the `coord_polar` coordinate system.  We'll use a pie chart to plot the proportion of characters from each home world, subsetted by the home worlds with the highest number of characters.  A more thorough description for creating pie charts with ggplot2 can be found here: https://www.r-graph-gallery.com/piechart-ggplot2.html.
  
```{r}
starwars %>% 
  group_by(homeworld) %>% 
  summarise(n = n()) %>% 
  filter(!is.na(homeworld)) %>% 
  arrange(desc(n)) %>% 
  filter(n > 2) %>% 
  ggplot(aes("", n, fill = homeworld)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  my_theme
```
  

2) _What's the difference between `geom_path()` and `geom_polygon()`?  What's the difference between `geom_path()` and `geom_line()`?_

The difference between `geom_path()` and `geom_polygon()` is the former draws lines between points in the order they appear in the data, but doesn't draw a final line to close and fill the path, unlike `geom_polygon()`.  When comparing `geom_path()` and `geom_line()`, the former draws lines between points as they appear in the data, while `geom_line()` connects points based on the order of the variable on the x-axis.  `geom_path()` may be used to show the path of movement of a vehicle on a map, for example, while `geom_line()` could be used to visualize time series data.

3) _What low-level geoms are used to draw `geom_smooth()`?  What about `geom_boxplot()` and `geom_violin()`?_

There are several low-level geoms that can be used to build more complex plots.  For example, it seems plausible that `geom_line()`, `geom_rect()`, and `geom_point()` can be combined to be used in `geom_boxplot()`: `geom_rect()` may draw the boundaries of the middle 50% of the data, `geom_line()` may be used to draw the median (the middle line in the boxplot) and the whiskers extending from the box, and `geom_point()` may be used to draw the outliers (if any exist).  In fact, when we look at the code for the `geom_boxplot()` function here (https://github.com/tidyverse/ggplot2/blob/master/R/geom-boxplot.r), we can see that the whiskers and box are rectangles and lines defined in the same way that you would use to draw them in a `geom_rect()` and `geom_line()` call, respectively.  In another case, `geom_smooth()` uses `geom_line()` to draw the smoothed mean response line, and `geom_violin()` relies on `geom_polygon()`.