---
title: 'Chapter 06: Maps'
author: "Stan Piotrowski"
date: "`r format(Sys.Date(), '%B %d %Y')`"
output:
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Setup 

This chapter focuses on drawing base maps and adding metadata from other sources using `geom_polygon()`, `geom_sf()`, and `geom_raster()`.

```{r}
# Libraries
suppressPackageStartupMessages(library(tidyverse))
library(ggtext)
library(ozmaps)
library(sf)
library(rmapshaper)

# Set ggplot2 theme
plot_theme <- theme(
  panel.background = element_rect(color = "black", fill = "white"),
  panel.grid = element_blank(),
  plot.title = element_markdown(),
  plot.subtitle = element_markdown(),
  plot.caption = element_markdown(),
  axis.title.x = element_markdown(),
  axis.title.y = element_markdown()
)
```

## Notes

Below are notes for the various subsections of this chapter:

### Simple features maps

Notes:
* Simple features objects generally contain a `geometry` field composed of a multipolygon list defining the shape of the object.
* The user doesn't need to explicity map the `geometry` aesthetic in a `geom_sf()` call- the function automatically looks for the field called `geometry` or if one doesn't exist, looks for a multipolygon field to map.
* Additional annotations can be mapped onto the same grob using standard ggplot2 syntax.

### Map projections

Notes:
* Geodetic datum refers to the set of assumptions about the shape of the Earth, which along with the map projection, dictates the location of a coordinate on a 2-D map.  Some of the most popular are the North American Datum (NAD83) and the World Geodetic System (WGS84).
* Map projections refer to how an ellipsoid shape like the Earth is mapped onto a 2-D plane (e.g., a plot).
* There are two types of map projections:
    1) Area-preserving projections- these ensure that regions of the same area are drawn with equal area on the map.
    2) Shape-preserving (conformal) projections- these ensure that the local shape of a region is preserved.
* The geodetic datum and the map projection together is referred to as the coordinate reference system (CRS), essentially a set of assumptions and rules to draw multipolygons onto a 2-D map.
* The `st_crs()` function can be used to inspect the CRS of a given sf object.
* The function `coord_sf()` ensures that all of the layers in a ggplot2 grob are plotted with the same projection, and using a similar behavior of mapping aesthetics to x and y coordinates uses the CRS of the `geometry` field in the data.
* In addition to retrieving the CRS from an object, `st_crs()` can be used to define the unambiguous strings for a given CRS using a compact syntax by defining the EPSG code. 
* For example, using `st_crs(ozmaps_states)` generates several lines of well-known text describing the CRS for the object, which is all summarized in the EPSG code 4283.  In fact, by typing `st_crs(4283)` we get the same well-known text string output.
* EPSG Geodetic Parameter Dataset is a collection of CRS and transformations that can be queried using a variety of methods (e.g., searching on a map, using strings, etc.).

### Raster maps

Notes:
* Raster data are in the form of images, unlike simple features format data, which are in the form of points, lines, and polygons.
